---
title: Calculate row sums and col means with tidyR
author: Sichong Peng
date: '2019-03-11'
slug: calculate-row-sums-and-col-means-with-tidyr
categories:
  - R
tags:
  - tidyverse
  - R
image:
  caption: ''
  focal_point: ''
knitr.table.format: markdown
---

Recently Someone asked a question on reddit:
[Adding a row and a column that are means for a set matrix](https://www.reddit.com/r/Rlanguage/comments/azwcs5/adding_a_row_and_a_column_that_are_means_for_a/)

This is a very simple problem that many of us learned how to do only a few hours in to basic R. But since I've been learning tidyverse package, I figured why not do it tidy-fashion?

Let's give it a try.

Here is the question:

>I have the following sequence:  
>```S<-seq(1,90, by=3)```  
>I make a matrix that is the following:  
>```matrix(S, nrow = 6, ncol = 5)```  
>Now I am trying to do the following:  
>I want to calculate the means of the columns of the matrix and add them as a new row under the columns.  
>Next I want to calculate the sum of the rows of the matrix and add them as a new column on the right of the matrix.  

Here is what their data looks like:
```{r}
data_matrix = matrix(seq(1, 90, by = 3), nrow = 6, ncol = 5)
data_matrix
```

They want to calculate the mean of each column and sum of each row.  
Now if you know some basic R this can very easily be achieved:
```{r, results='asis'}
data_df = data.frame(data_matrix)
data_df[,ncol(data_df)+1] = rowSums(data_df)
data_df[nrow(data_df)+1,] = colMeans(data_df)
data_df
```

But how can we do this "tidy" way? Turns out this is more complicated than I originally thought.

First let's breath some context into our data so we don't get too bored:  
Let's say we had an exam in a class of 50 students. The exam consists of 5 questions. TAs recorded the score of each question for each student on an excel sheet. The spreadsheet data has 5 columns, each for a question and 50 rows, each for a student. Now we want to calculate:  
1. Total score for each student (row sums)  
2. Class average for each question (col means)  

Let's first creat this "spreadsheet"
```{r}
scores_df = data.frame(matrix(sample(1:20,250, replace = TRUE), nrow = 50, ncol = 5))
head(scores_df)
```

Here we have a data frame with 5 colmuns (5 questions) and 50 rows (50 students).  
Now the next step is to for tidy data: key-value pairs.  
First we need to identify the keys: in our dataset, each student-question pair uniquely identifies a value (score). So we can gather our date like this:
```{r warning=FALSE, message=FALSE}
library(tidyverse)
scores_tibble = gather(rownames_to_column(scores_df, var = "student"), question, score, -student)
head(scores_tibble)
```
Here I did not ```unite()``` ```student``` and ```question``` into a single column so that later I can group by either student or question to calculate mean and sum.  
Now to calculate question means:
```{r}
summarise(group_by(scores_tibble, question), mean = mean(score))
```

Student sum scores:
```{r}
summarise(group_by(scores_tibble, student), total = sum(score))
```

Let's get a single table with all information
```{r}
scores_complete = group_by(scores_tibble, question) %>%
  mutate(question_average = mean(score)) %>%
    group_by(student) %>%
      mutate(student_total = sum(score))
scores_complete
```

With a tidy table we can easily calculate stats we want and add them to the original data frame:
```{r}
scores_df_complete = spread(scores_tibble, question, score) %>%
  left_join(summarise(group_by(scores_tibble, student), student_total = sum(score)), by = "student") %>%
    bind_rows(spread(summarise(group_by(scores_tibble, question), question_mean = mean(score)), question, question_mean))
head(scores_df_complete)
tail(scores_df_complete)
```

